Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-01 code/MSstatsPTM_Simulation_Analysis.Rmd
Fixed all plots

png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)

models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
                                               "Limma no adjustment", "T-Test", "T-Test no adjustment")),
                  Replicates = as.factor(rep)) %>%
  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
  scale_colour_manual(values=cbPalette_first) +
  scale_size_manual(values = c(3,5,7,9)) + 
  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12), 
         axis.text.y = element_text(size = 14), 
         legend.text=element_text(size=14),
         axis.title.y = element_text(size = 22),
         axis.title.x = element_text(size = 22),
         title = element_text(size = 22),
         strip.text = element_text(size = 16)) + 
  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR") + 
     guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
              size = guide_legend(order = 2, reverse = TRUE))

>>>>

png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
                                               "Limma no adjustment", "T-Test", "T-Test no adjustment")),
                  ) %>%
  filter(rep == 5) %>% 
  ggplot() + geom_point(aes(x = Model, y = fpr, color = Model), size = 9) + #, size = Replicates
  scale_colour_manual(values=cbPalette_first) +
  # scale_size_manual(values = c(3,5,7,9)) + 
  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
  theme_bw() +
  ylim(0.0, 0.45) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 12), 
         axis.text.y = element_text(size = 14), 
         legend.text=element_text(size=14),
         axis.title.y = element_text(size = 22),
         axis.title.x = element_text(size = 22),
         title = element_text(size = 22),
         strip.text = element_text(size = 16)) + 
  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR") + 
     guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
              size = guide_legend(order = 2, reverse = TRUE))

### 2022-02-01 code/Spike-in_Analysis.Rmd
Fixed all plots

@@

@@

@@

@@

spike <- rbindlist(list(unadj_spike_msstats, adj_spike_msstats, unadj_spike_ttest,
                        adj_spike_ttest, unadj_spike_limma, adj_spike_limma), fill = TRUE)
spike <- as.data.table(spike)
spike[spike$Model == "MSstatsPTM", Model := "Proposed"]
#plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']]

png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
                 Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%  ggplot() + 
  geom_boxplot(aes(x = Label, y = log2FC, fill = Model)) + 
  geom_point(aes(y = trueLog2FC, x = Label), color = "red", shape = 4, size = 4, stroke = 2) + 
      scale_fill_manual(values=cbPalette) +
 facet_wrap(.~Adjustment) + 
    theme_bw() + 
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
       axis.text.y = element_text(size = 16), 
       legend.text=element_text(size=18),
       axis.title.y = element_text(size = 18),
       axis.title.x = element_text(size = 18),
       title = element_text(size = 20),
       strip.text = element_text(size = 18),
       legend.position = "bottom") + 
  labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
dev.off()
```

>>>>

proposed_plot_df <- plot_df

@@

#### Calculating background peptide percentages

```{r, echo=FALSE}

light_pep <- model_df$PTM.Model %>% filter(labeling == "light")

unadj_sig_light <- model_df$PTM.Model %>% filter(labeling == "light" & adj.pvalue < .05)
adj_sig_light <- model_df$ADJUSTED.Model %>% filter(labeling == "light" & adj.pvalue < .05)

nrow(unadj_sig_light) / nrow(light_pep)
nrow(adj_sig_light) / nrow(light_pep)

@@

limma_plot_df <- plot_df

@@

#### Create combo plot of proposed and limma

``` {r, echo = FALSE}

proposed_plot_df$Model <- "Proposed Method"
limma_plot_df$Model <- "Limma"
limma_plot_df$Protein <- limma_plot_df$PTM
limma_plot_df$log2FC <- limma_plot_df$Log2FC

combo_plot_df <- rbind(proposed_plot_df[, c("Protein", "Label", "log2FC", 
                                            "adj.pvalue", "labeling", 
                                            "trueLog2FC", "contrast_truth", 
                                            "protadj", "Model")], 
                       limma_plot_df[, c("Protein", "Label", "log2FC", 
                                            "adj.pvalue", "labeling", 
                                            "trueLog2FC", "contrast_truth", 
                                            "protadj", "Model")])
combo_plot_df$Model <- factor(combo_plot_df$Model, levels=c('Proposed Method','Limma'))

png("../main/images/spike_in_volcano.png", width = 750, height = 500)
combo_plot_df %>% arrange(Model, labeling) %>% 
  filter(abs(log2FC) < Inf & Label %in% c("mix2-mix3", "mix2-mix4")) %>%
  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
  scale_color_manual(values = c("#999999", "#D55E00")) + 
  scale_alpha_manual(values = c(0.2, 1)) + 
  #scale_size_manual(values = c(2, 1)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0, linetype = "dashed") + 
  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
  facet_grid(protadj ~ Model + contrast_truth) + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 16), 
       axis.text.y = element_text(size = 16), 
       legend.text=element_text(size=16),
       axis.title.y = element_text(size = 18, vjust=2),
       axis.title.x = element_text(size = 18),
       title = element_text(size = 20),
       strip.text = element_text(size = 15),
       legend.title =  element_blank(),
       legend.position = "bottom",
       legend.box.margin=margin(-10,-10,-10,-10)) + 
  scale_y_continuous(limits = c(0,3)) + 
  guides(colour = guide_legend(override.aes = list(size=5))) +
  labs(title = "SpikeIn VolcanoPlot", x= "Log2FC")
dev.off()

@@

spike <- rbindlist(list(unadj_spike_msstats, adj_spike_msstats, unadj_spike_ttest,
                        adj_spike_ttest, unadj_spike_limma, adj_spike_limma), fill = TRUE)
spike <- as.data.table(spike)
spike[spike$Model == "MSstatsPTM", Model := "Proposed"]
#plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']]
spike$`color_label` <- "True LogFC"

png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
      Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%
  filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
  ggplot() + 
  geom_boxplot(aes(x = Label, y = log2FC, fill = Model)) + 
  geom_point(aes(y = trueLog2FC, x = Label, color = color_label), 
             shape = 4, size = 4, stroke = 2) + 
  scale_fill_manual(values=cbPalette) +
  scale_color_manual(name = "", values="red", label="True LogFC") + 
 facet_wrap(.~Adjustment) + 
    theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 16), 
       axis.text.y = element_text(size = 16), 
       legend.text=element_text(size=18),
       axis.title.y = element_text(size = 18),
       axis.title.x = element_text(size = 18),
       title = element_text(size = 20),
       strip.text = element_text(size = 18),
       legend.position = "bottom") + 
  labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
dev.off()
```

### 2022-02-01 main/MSstatsPTM_Main.tex
Fixed all plots

There are two main biological questions that are generally relevant in PTM analysis. First, some researchers run mass-spectrometry based proteomic experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, correcting for confounding with the total protein is relevant. These two questions are both important areas of study, and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address the convolution with the total protein.

@@

Here simulations were generated mimicking a Label-free experiment, with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with different numbers of replicates and conditions. In all simulations 1000 modified peptides were generated, half of which were true positives and the other half true negatives. Half of the true positives and true negatives were generated convoluted with global protein abundance. Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{Dataset 1 : Computer simulation 1 - Label-free. All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performing method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test. While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
\label{fig:sim_boxplot}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{}
\label{fig:spikein_limma_volcano}
\end{subfigure}
\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free. A) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjustment the spike in peptides are more in line with expectation. Additionally, the background peptides showed many false positives before adjustment, but decrease considerably after adjustment. B) Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level. However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.}
\label{fig:spikein_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.5\linewidth}
\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
\caption{}
\label{fig:data4_venn_diagram}
\end{subfigure}
\begin{subfigure}[c]{0.95\linewidth}
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{}
\label{fig:data4_profile_plot}
\end{subfigure}
\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT. A) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein. In contrast, peptides that became significant after adjustment saw their true abundance change masked by underlying changes in the unmodified protein. B) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.}
\label{fig:data4_plots}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment. Again more PTMs became insignificant after adjustment then became significant.}
\label{fig:shig_venn_d}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend. Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicates the movement is driven by changes in global protein.}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$. In this case the modification and global protein trend in different directions. Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change. In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance. Additionally this profile plot shows the large difference in available features between modifications and global protein.}
 \end{subfigure}
\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
\label{fig:shigella_profile}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm}
	\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. Here many more PTMs became insignificant then became significant after adjustment. This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm_matching_only}
	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed. In this case we see significantly less peptides become insignificant after adjustment. This highlights the need for a global profiling run if protein adjustment is going to performed.}
 \end{subfigure}
 \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
\label{fig:usp30_ven_d}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
\caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$. The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE. This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
\label{fig:usp30_profile}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.725\textwidth]{images/same_var_power}
	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides. Predictably when the replicates are high for both modified and unmodified peptides the power is much higher. Conversely at low replicates for each the power is much lower. The interesting part of this chart is when the replicates are different between runs. With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{images/high_ptm_var_power}
	\caption{In this chart the variance for the PTM is higher than the unmodified protein. The PTM variance is .2, while the unmodified protein variance is .1. With equal replicates the results are the same as above, more replicates equals more power. When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
 \end{subfigure}
 \caption{Power analysis of experiments with differing variances.}
\label{fig:power_sd_combo}
\end{figure}

>>>>

Generally, there are two main biological questions that are relevant in PTM analysis. First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant. These two questions are both important areas of study and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address confounding with the total protein.

@@

Here simulations were generated mimicking a Label-free experiment with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with different numbers of replicates and conditions. In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives. Half of the true positives and true negatives were generated convoluted with global protein abundance. Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. These results were then compared to the ground truth to determine TP and TN. We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

For biological experiments where ground truth was not known, we again adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\end{subfigure}
\caption{Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach is apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperforms Limma and t-test in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_limma_volcano}
\end{subfigure}
\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free. a) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjustment the spike in peptides are more in line with expectation. Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level. However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. b) say something about this one.}
\label{fig:spikein_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.5\linewidth}
\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
\caption{}
\label{fig:data4_venn_diagram}
\end{subfigure}
\begin{subfigure}[c]{0.95\linewidth}
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{}
\label{fig:data4_profile_plot}
\end{subfigure}
\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT. a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein. In contrast, peptides that became significant after adjustment saw their true abundance change masked by underlying changes in the unmodified protein. b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.}
\label{fig:data4_plots}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
\includegraphics[height=.5\textwidth]{images/shig_venn_diagramm.png}
\caption{}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{}
	 \end{subfigure}
\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment. Again more PTMs became insignificant after adjustment then became significant. Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$. When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend. Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicates the movement is driven by changes in global protein that is only accounted for in the model when adjusting for global protein abundance change.}
\label{fig:shigella_plots}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm}
	\caption{}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm_matching_only}
	\caption{}
 \end{subfigure}
 \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. a) The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. Here many more PTMs became insignificant then became significant after adjustment. This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins. b) Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed. In this case we see significantly less peptides become insignificant after adjustment. This highlights the need for a global profiling run if protein adjustment is going to performed.}
\label{fig:usp30_ven_d}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{images/same_var_power}
	\caption{}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{images/high_ptm_var_power}
	\caption{}
 \end{subfigure}
 \caption{Power analysis of experiments with differing variances. a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides. Predictably when the replicates are high for both modified and unmodified peptides the power is much higher. Conversely at low replicates for each the power is much lower. The interesting part of this chart is when the replicates are different between runs. With equal variance, it does not matter if the PTM replicates or protein replicates are higher. b) In this chart the variance for the PTM is higher than the unmodified protein. The PTM variance is .2, while the unmodified protein variance is .1. With equal replicates the results are the same as above, more replicates equals more power. When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
\label{fig:power_sd_combo}
\end{figure}

### 2022-02-01 supplementary/ptm_sm.tex
Fixed all plots

In experiments of complex designs, multiple inter-related conditions are often compared together. The proposed approach and Limma leverage measurements in all conditions for the inference of the underlying abundance, whereas $t$-test uses measurements from the two conditions being compared. To highlight this distinction, multiple conditions of data were generated. Two simulations were generated based on the following parameters:

@@

A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting. Heavy-labeled KGG modified peptides were used as spike-in peptides. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The KGG enriched dataset included the spike-in peptides, as well as modified and unmodified human lysate. The LC-MS dataset included only unmodified peptides. The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases, whereas the human lysate modified peptides were treated as controls and are not expected to be differential in any comparison.

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$. In this case the modification and global protein trend in different directions. Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change. In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance. Additionally this profile plot shows the large difference in available features between modified peptides and global proteins.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
	\caption{Volcano plots of WT\_Early vs WT\_Uninfect both before and after protein adjustment. The $KI67\_MOUSE|E9PVX6$ modification is highlighted. Before adjustment the modification has a small positive fold change and insignificant adjusted pvalue. After adjustment the fold change increases and the adjusted pvalue is now significant.}
 \end{subfigure}
 \caption{Summary plots for modification of protein KI67 at site T215.}
\label{fig:Diff_Shigella_PTM}
\end{figure}

>>>>

To compare performance across methods, we first calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To determine TP and TN we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. These results were then compared to the ground truth to determine TP and TN. We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

In experiments of complex designs, multiple inter-related conditions are often compared together. The proposed approach and Limma leverage measurements in all conditions for the inference of the underlying abundance, whereas $t$-test uses measurements from the two conditions being compared. To highlight this distinction, multiple conditions of data were generated. While multiple conditions were generated, all comparisons were still made with the same fold change between conditions ($0.75$). This ensured that all peptides were differentially abundant by the same amount and the only difference in TP/FP/TN/FN calculations was using more conditions in the modeling.

@@

A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting. Heavy-labeled KGG modified peptides were used as spike-in peptides. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The KGG enriched dataset included the spike-in peptides, as well as modified and unmodified human lysate. The LC-MS dataset included only unmodified peptides. The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases. Additionally, background E coli lysate was used to normalize total protein levels prior to the enrichment and global protein profiling. The background lysate was treated as controls and are not expected to be differential in any comparison.

\begin{figure}[h!]
\centering
\includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
\caption{An overview of the SpikeIn benchmark experimental design. The corresponding levels of SpikeIn peptide, human lysate, and background lysate for each mixture are shown. The SpikeIn peptide levels are adjusted for the human lysate levels. For example, the true SpikeIn abundance of mixture 1 is entirely accounted for by human lysate, whereas mix 2 has double the amount human lysate compared to SpikeIn abundance. When comparing mix 1 vs mix 2 we would expect to see a log fold change of -1. In this way we know the expected log fold change between mixture comparisons. \label{fig:spikein_overview}}
\end{figure}

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$. In this case the modification and global protein appear to show very small to no differences between conditions, however after adjusting for change in global protein abundance, the modification is statistically significant. Additionally, this profile plot shows the large difference in available features between modified peptides and global proteins.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
	\caption{The volcano plot of the WT\_Uninfect and WT\_Early comparison show the specifics of the adjustment. The modification looks to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$. While both exhibit small changes, when combined we can see a log fold change of $.804$ and adjusted p-value of $.0156$. In this case we can see the strength of the proposed method, showing noteworthy modifications which would be missed otherwise.}
 \end{subfigure}
 \caption{Summary plots for modification of protein KI67 at site T215.}
\label{fig:Diff_Shigella_PTM}
\end{figure}
